---
title: "In-Class_Ex1"
editor: visual
date: 2023-11-18
date-format: long
date-modified: "last-modified"
---

## Getting started

The code chunk below load the following packages:

-   tmap for thematic mapping

-   sf for geospatial data handling

-   tidyverse for non-spatial data handling

    \*\*[Pacman](https://cran.r-project.org/web/packages/pacman/readme/README.html)\*\*

```{r}
pacman::p_load(tmap, sf, tidyverse)
```

## Preparing the data

## Importing the 00 data

Firstly, we will import the "*Passenger Volume by Origin Destination Bus Stops"* data set downloaded from LTA Datamall by using 'read.csv()' of *readr* package

```{r}
obus <- read.csv("data/aspatial/origin_destination_bus_202308.csv")
```

```{r}
obus$ORIGIN_PT_CODE <- as.factor(obus$ORIGIN_PT_CODE)
obus$DESTINATION_PT_CODE <- as.factor(obus$DESTINATION_PT_CODE)
```

```{r}
oritrip_7_9 <- obus %>% filter(DAY_TYPE == "WEEKDAY") %>%   
  filter(TIME_PER_HOUR >= 7 & TIME_PER_HOUR <=9) %>%   
  group_by(ORIGIN_PT_CODE)%>%   summarise(TRIPS = sum(TOTAL_TRIPS))
```

Import bus stop

```{r}
busstop <- st_read(dsn = "data/geospatial", layer = "BusStop") %>%   st_transform(crs = 3414)
```

```{r}
mpsz <- st_read(dsn = "data/geospatial", layer = "MPSZ-2019") %>%   st_transform(crs = 3414)
```

```{r}
mpsz
```
