---
title: "In Class Exercise 2: Emerging Hot Spot Analysis: sfdep methods"
author: "Chai Zhixuan"
editor: visual
date: 2023-11-25
date-format: long
date-modified: "last-modified"
---

# **Getting Started**

Five R packages will be used in this in-class exercise.

Please note that [plotly](https://plotly.com/r/) is used in this hands-on exercise:

```{r}
pacman::p_load(tmap, sf, tidyverse, knitr, sfdep, plotly)

```

To take note on the learning points for this hands-on exercise

Emerging Hot Spot Analysis (EHSA) is a spatio-temporal analysis method for revealing and describing how hot spot and cold spot areas evolve over time. The analysis consist of four main steps:

-   Building a space-time cube,

-   Calculating Getis-Ord local Gi\* statistic for each bin by using an FDR correction,

-   Evaluating these hot and cold spot trends by using Mann-Kendall trend test,

-   Categorising each study area location by referring to the resultant trend z-score and p-value for each location with data, and with the hot spot z-score and p-value for each bin.

# The Data

Same data will be used, except that apatial will be Hunan_GDPPC, an attribute data set in csv format instead.

## Import geospatial data

```{r}
hunan <- st_read(dsn = "data/geospatial",
                   layer = "Hunan")
```

## Import apatial data

```{r}
GDPPC <- read_csv("data/aspatial/Hunan_GDPPC.csv")
```

## Creating a Time Series Cube

In the code chunk below, [`spacetime()`](https://sfdep.josiahparry.com/reference/spacetime.html) of sfdep is used to create an spacetime cube.

```{r}
GDPPC_st <- spacetime(GDPPC, hunan,
                      .loc_col = "County",
                      .time_col = "Year")
```

Next, `is_spacetime_cube()` of sfdep package will be used to varify if GDPPC_st is indeed an space-time cube object.

```{r}
is_spacetime_cube(GDPPC_st)
```

To note: The **TRUE** return confirms that *GDPPC_st* object is indeed an time-space cube.

If it is ***FALSE***, it will have complications

## Computing GI\*

The code chunk below will be used to identify neighbors and to derive an inverse distance weights.

```{r}
GDPPC_nb <- GDPPC_st %>%
  activate("geometry") %>%
  mutate(nb = include_self(st_contiguity(geometry)),
         wt = st_inverse_distance(nb, geometry,
                                  scale = 1,
                                  alpha = 1),
         .before = 1) %>%
  set_nbs("nb") %>%
  set_wts("wt")
```

Note that this dataset now has neighbors and weights for each time-slice.

```{r}
head(GDPPC_nb)
```

We can use these new columns to manually calculate the local Gi\* for each location. We can do this by grouping by *Year* and using `local_gstar_perm()` of sfdep package. After which, we `use unnest()` to unnest *gi_star* column of the newly created *gi_starts* data.frame.

```{r}
gi_stars <- GDPPC_nb %>% 
  group_by(Year) %>% 
  mutate(gi_star = local_gstar_perm(
    GDPPC, nb, wt)) %>% 
  tidyr::unnest(gi_star)
```

## **Mann-Kendall Test**

With these Gi\* measures we can then evaluate each location for a trend using the Mann-Kendall test. The code chunk below uses Changsha county.

```{r}
cbg <- gi_stars %>% 
  ungroup() %>% 
  filter(County == "Changsha") |> 
  select(County, Year, gi_star)
```

Next, we plot the result by using ggplot2 functions.

```{r}
ggplot(data = cbg, 
       aes(x = Year, 
           y = gi_star)) +
  geom_line() +
  theme_light()
```

To take note: We can also create an interactive plot by using `ggplotly()` of **plotly** package.

```{r}
p <- ggplot(data = cbg, 
       aes(x = Year, 
           y = gi_star)) +
  geom_line() +
  theme_light()

ggplotly(p)
```

```{r}
cbg %>%
  summarise(mk = list(
    unclass(
      Kendall::MannKendall(gi_star)))) %>% 
  tidyr::unnest_wider(mk)
```

In the above result, sl is the p-value. This result tells us that there is a slight upward but insignificant trend.

We can replicate this for each location by using `group_by()` of dplyr package.

```{r}
ehsa <- gi_stars %>%
  group_by(County) %>%
  summarise(mk = list(
    unclass(
      Kendall::MannKendall(gi_star)))) %>%
  tidyr::unnest_wider(mk)
```

## **Arrange to show significant emerging hot/cold spots**

```{r}
emerging <- ehsa %>% 
  arrange(sl, abs(tau)) %>% 
  slice(1:5)
```

## **Performing Emerging Hotspot Analysis**

Lastly, we will perform EHSA analysis by using [`emerging_hotspot_analysis()`](https://sfdep.josiahparry.com/reference/emerging_hotspot_analysis.html) of sfdep package.

To note: It takes a spacetime object x (i.e. GDPPC_st), and the quoted name of the variable of interest (i.e. GDPPC) for .var argument. The k argument is used to specify the number of time lags which is set to 1 by default. Lastly, nsim map numbers of simulation to be performed.

```{r}
ehsa <- emerging_hotspot_analysis(
  x = GDPPC_st, 
  .var = "GDPPC", 
  k = 1, 
  nsim = 99
)
```

## **Visualising the distribution of EHSA classes**

In the code chunk below, ggplot2 functions ised used to reveal the distribution of EHSA classes as a bar chart.

```{r}
ggplot(data = ehsa,
       aes(x = classification)) +
  geom_bar()
```

## **Visualising EHSA**

In this section, I will learn how to visualise the geographic distribution EHSA classes. However, before I can do so, I need to join both *hunan* and *ehsa* together by using the code chunk below.

```{r}
hunan_ehsa <- hunan %>%
  left_join(ehsa,
            by = join_by(County == location))
```

Next, tmap functions will be used to plot a categorical choropleth map by using the code chunk below.

```{r}
ehsa_sig <- hunan_ehsa  %>%
  filter(p_value < 0.05)
tmap_mode("plot")
tm_shape(hunan_ehsa) +
  tm_polygons() +
  tm_borders(alpha = 0.5) +
tm_shape(ehsa_sig) +
  tm_fill("classification") + 
  tm_borders(alpha = 0.4)
```
