---
title: "In Class Exercise 2: Spatial Weights - sfdep methods and Local Measures of Spatial Association "
author: "Chai Zhixuan"
editor: visual
date: 2023-11-25
date-format: long
date-modified: "last-modified"
---

# **Getting Started**

Four R packages will be used in this in-class exercise.

To understand sfdep better, click on this [link](https://sfdep.josiahparry.com/).

\*Note: sfdep is not covered in the hands on exercise and will be used in the take home exercise

```{r}
pacman::p_load(tmap, sf, tidyverse, knitr, sfdep)
```

# The Data

For the purpose of this in-class exercise, the Hunan data sets will be used. There are two data sets in this use case, they are:

-   Hunan, a geospatial data set in ESRI shapefile format, and

-   Hunan_2012, an attribute data set in csv format.

## Importing apatial data

```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")
```

## Importing geospatial data

```{r}
hunan <- st_read(dsn = "data/geospatial",
                   layer = "Hunan")
```

```{r}
head(hunan)
```

## **Performing relational join**

Two datasets are joined and select columns are retained in the resulting table.

Do an intermediate step to see the resulting table and choose the columns to retain. "Geometry" column will be automatically bring it in.

Refer to the link [here](https://dplyr.tidyverse.org/reference/mutate-joins.html) for the information for left join in dplyr.

Code is as shown below:

```{r}
hunan <- left_join(hunan,hunan2012)%>%
  select(1:4, 7, 15)
```

## **Plotting a choropleth map**

The choropleth should look similar to ther figure below.

```{r}
tmap_mode("plot")
tm_shape(hunan) +
  tm_fill("GDPPC", 
          style = "quantile", 
          palette = "Blues",
          title = "GDPPC") +
  tm_borders(alpha = 0.5) +
  tm_layout(main.title = "Distribution of GDP per capita by district, Hunan Province",
            main.title.position = "center",
            main.title.size = 1.2,
            legend.height = 0.45, 
            legend.width = 0.35,
            frame = TRUE) +
  tm_compass(type="8star", size = 2) +
  tm_scale_bar() +
  tm_grid(alpha =0.2)
```

## Deriving contiguity weights: Queen's method

Below shows the method to compute the contiguity weights by using [`st_weights()`](https://sfdep.josiahparry.com/reference/st_weights.html) of **sfdep** package.

In the code chunk below, queen method is used to derive the contiguity weights.

```{r}
wm_q <- hunan%>%
  mutate(nb = st_contiguity(geometry),
         wt = st_weights(nb,
                         style = "W"),
         .before =1)
```

What we can learn from `st_weights()`. It provides tree arguments, they are:

-   *nb*: A neighbor list object as created by st_neighbors().

-   *style*: Default \"W\" for row standardized weights. This value can also be \"B\", \"C\", \"U\", \"minmax\", and \"S\". B is the basic binary coding, W is row standardised (sums over all links to n), C is globally standardised (sums over all links to n), U is equal to C divided by the number of neighbours (sums over all links to unity), while S is the variance-stabilizing coding scheme proposed by Tiefelsdorf et al. 1999, p. 167-168 (sums over all links to n).

-   *allow_zero*: If TRUE, assigns zero as lagged value to zone without neighbors.

```{r}
wm_q
```

## Computing local Moran's I

In this section, I learn how to compute Local Moran\'s I of GDPPC at county level by using [`local_moran()`](https://sfdep.josiahparry.com/reference/local_moran.html) of sfdep package:

```{r}
lisa <- wm_q %>%
  mutate(local_moran = local_moran(
    GDPPC, nb, wt, nsim = 99),
    .before =1) %>%
  unnest(local_moran)
```

To take note:

The output of `local_moran()` is a sf data.frame containing the columns ii, eii, var_ii, z_ii, p_ii, p_ii_sim, and p_folded_sim.

-   ii: local moran statistic

-   eii: expectation of local moran statistic; for localmoran_permthe permutation sample means

-   var_ii: variance of local moran statistic; for localmoran_permthe permutation sample standard deviations

-   z_ii: standard deviate of local moran statistic; for localmoran_perm based on permutation sample means and standard deviations p_ii: p-value of local moran statistic using pnorm(); for localmoran_perm using standard deviatse based on permutation sample means and standard deviations p_ii_sim: For `localmoran_perm()`, `rank()` and `punif()` of observed statistic rank for \[0, 1\] p-values using `alternative=` -p_folded_sim: the simulation folded \[0, 0.5\] range ranked p-value (based on https://github.com/pysal/esda/blob/4a63e0b5df1e754b17b5f1205b cadcbecc5e061/esda/crand.py#L211-L213)

-   skewness: For `localmoran_perm`, the output of e1071::skewness() for the permutation samples underlying the standard deviates

-   kurtosis: For `localmoran_perm`, the output of e1071::kurtosis() for the permutation samples underlying the standard deviates.

## **Visualising local Moran\'s I**

In this code chunk below, tmap functions are used prepare a choropleth map by using value in the *ii* field.

```{r}
tmap_mode("plot")
tm_shape(lisa) +
  tm_fill("ii") + 
  tm_borders(alpha = 0.5) +
  tm_view(set.zoom.limits = c(6,8)) +
  tm_layout(main.title = "local Moran's I of GDPPC",
            main.title.size = 0.8)
```

## **Visualising p-value of local Moran\'s I**

In the code chunk below, tmap functions are used prepare a choropleth map by using value in the *p_ii_sim* field.

```{r}
tmap_mode("plot")
tm_shape(lisa) +
  tm_fill("p_ii_sim") + 
  tm_borders(alpha = 0.5) +
   tm_layout(main.title = "p-value of local Moran's I",
            main.title.size = 0.8)
```

## **Visuaising local Moran\'s I and p-value**

For effective comparison, it will be better for us to plot both maps next to each other as shown below.

```{r}
tmap_mode("plot")
map1 <- tm_shape(lisa) +
  tm_fill("ii") + 
  tm_borders(alpha = 0.5) +
  tm_view(set.zoom.limits = c(6,8)) +
  tm_layout(main.title = "local Moran's I of GDPPC",
            main.title.size = 0.8)

map2 <- tm_shape(lisa) +
  tm_fill("p_ii",
          breaks = c(0, 0.001, 0.01, 0.05, 1),
              labels = c("0.001", "0.01", "0.05", "Not sig")) + 
  tm_borders(alpha = 0.5) +
  tm_layout(main.title = "p-value of local Moran's I",
            main.title.size = 0.8)

tmap_arrange(map1, map2, ncol = 2)
```

## **Visualising LISA map**

LISA map is a categorical map showing outliers and clusters. There are two types of outliers namely: High-Low and Low-High outliers. Likewise, there are two type of clusters namely: High-High and Low-Low cluaters. In fact, LISA map is an interpreted map by combining local Moran\'s I of geographical areas and their respective p-values.

In lisa sf data.frame, we can find three fields contain the LISA categories. They are *mean*, *median* and *pysal*. In general, classification in *mean* will be used as shown in the code chunk below.

```{r}
lisa_sig <- lisa  %>%
  filter(p_ii < 0.05)
tmap_mode("plot")
tm_shape(lisa) +
  tm_polygons() +
  tm_borders(alpha = 0.5) +
tm_shape(lisa_sig) +
  tm_fill("mean") + 
  tm_borders(alpha = 0.4)
```

## **Hot Spot and Cold Spot Area Analysis (HCSA)**

HCSA uses spatial weights to identify locations of statistically significant hot spots and cold spots in an spatially weighted attribute that are in proximity to one another based on a calculated distance. The analysis groups features when similar high (hot) or low (cold) values are found in a cluster. The polygon features usually represent administration boundaries or a custom grid structure.

```{r}
wm_idw <- hunan %>%
  mutate(nb = st_contiguity(geometry),
         wts = st_inverse_distance(nb, geometry,
                                   scale = 1,
                                   alpha = 1),
         .before = 1)
```

Next, [`local_gstar_perm()`](https://sfdep.josiahparry.com/reference/local_gstar) of sfdep package will be used to compute local Gi\* statistics as shown in the code chunk below.

```{r}
HCSA <- wm_idw %>% 
  mutate(local_Gi = local_gstar_perm(
    GDPPC, nb, wt, nsim = 99),
         .before = 1) %>%
  unnest(local_Gi)
HCSA
```

## **Visualising Gi\***

```{r}
tmap_mode("plot")
tm_shape(HCSA) +
  tm_fill("gi_star") + 
  tm_borders(alpha = 0.5) +
  tm_view(set.zoom.limits = c(6,8))
```

## **Visualising p-value of HCSA**

```{r}
tmap_mode("plot")
tm_shape(HCSA) +
  tm_fill("p_sim") + 
  tm_borders(alpha = 0.5)
```

## **Visuaising local HCSA**

```{r}
tmap_mode("plot")
map1 <- tm_shape(HCSA) +
  tm_fill("gi_star") + 
  tm_borders(alpha = 0.5) +
  tm_view(set.zoom.limits = c(6,8)) +
  tm_layout(main.title = "Gi* of GDPPC",
            main.title.size = 0.8)

map2 <- tm_shape(HCSA) +
  tm_fill("p_value",
          breaks = c(0, 0.001, 0.01, 0.05, 1),
          labels = c("0.001", "0.01", "0.05", "Not sig")) + 
  tm_borders(alpha = 0.5) +
  tm_layout(main.title = "p-value of Gi*",
            main.title.size = 0.8)

tmap_arrange(map1, map2, ncol = 2)
```

## **Visualising hot spot and cold spot areas**

We plot the significant (i.e. p-values less than 0.05) hot spot and cold spot areas by using appropriate tmap functions as shown below.

```{r}
HCSA_sig <- HCSA  %>%
  filter(p_sim < 0.05)
tmap_mode("plot")
tm_shape(HCSA) +
  tm_polygons() +
  tm_borders(alpha = 0.5) +
tm_shape(HCSA_sig) +
  tm_fill("gi_star") + 
  tm_borders(alpha = 0.4)
```

To notes (from hands on): Figure above reveals that there is one hot spot area and two cold spot areas. Interestingly, the hot spot areas coincide with the High-high cluster identifies by using local Moran\'s I method in the earlier sub-section
